A high-level programming framework to support parallel processing of communication intensive \\ stream applications

### 2013-03-20 paper.tex
Kevin's changes

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

\begin{abstract}
%\boldmath
the abstract goes here.

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Motivation}
In the continued race for improvement of multi core and many core architectures, computer networks are becoming a fundamental area for performance bottleneck especially in systems which are used for distributed processing of parallel applications. In order to handle the requirements posed by these applications the interconnects are becoming more and more complex. Such systems utilize a wide range of custom interconnects ranging from infiniband, myrinet to the more readily available ones such as 1Gbe and 10Gbe Ethernet standards and sophisticated routers and switches to handle the added challenge in routing data. While a small group of the specialized systems employ custom routers and topology configurations the majority of them especially those that use 1Gbe or 10Gbe use the commercial routers. In addition the growing complexity of routing protocols has resulted in advanced switches such as OpenVSwitch \cite{openvswitch}, etc., has resulted in these routers being upgraded frequently to keep up with the growing requirements. Software routers such as click modular router have been gaining popularity due to their flexible nature and cheaper alternative to the more expensive commercial routers and are widely used to prototype newer standards. In using software routers in the above scenario it would be possible to not only benefit from the newer developments in switching and routing protocols but it would require little effort to implement it in existing systems. One might argue that it would be a challenge to handle such complex workloads in software routers but with the presence of multicore and many core architectures and stream processors it is feasible to parallelize the packet processing workloads to improve the overall performance of the system. Several research projects are working on improving the performance by parallelizing the packet processing tasks on these architectures \cite{routebricks} \cite{packetshader}.

In a computing cluster which is used in the execution of parallel applications it is common to dedicate a core or processor in each node for operating system related tasks in order to minimize its effects on the application. Similarly, it should also be possible to utilize an additional core or processor to perform packet processing operations. While this is would be a challenging scenario it does provide the added benefits of a cheaper alternative to commercial routers. Also in bringing the network system into a software controlled domain it should be possible to perform additional optimizations from the application such as prioritized data flow, intelligent routing, etc., that would otherwise be difficult to implement due to the abstraction of the network system.

The focus of this paper is the development of a parallel processing engine that could be integrated with the application and can perform both the actual computations and the packet processing related tasks in parallel. The entire system can be divided into key parts as follows,

@@

\item A suitable methodology by which we can access the data in the userspace without interacting with the operating system. To do this (1)We need to read the data from the Network Interface Controller (NIC) before the OS and (2)Using this, establish a efficient manner for sending and receiving the data.(Section \ref{sec:userspace-packet})

\item Given a graph that represents the workload to be done on the packet flow, we divide it into tasks that can be executed in parallel and determine the regions that could be used to extract data and task parallelism.(Section \ref{sec:workload})

\item The parallel processing engine that utilizes the above and does the actual computations on the multi core or many core processor. It also handles load balancing and scheduling of the work across the available computing resources. (Section \ref{sec:ppe})

@@

We concentrate more on the framework that can integrate the packet processing and application tasks into a parallel processing engine and the extraction of parallelism from the packet processing related tasks to improve the overall performance. Given an application graph and its partition consisting of parallel tasks we the combine it with the packet processing operations. In addition, Section \ref{sec:} we show the challeneges involved in using a system consisting of threads and process that are used to actually perform the work. The extent to which we extract parallelism ranges not only to the packet tasks but to the low level access of data I/O to the network interface hardware through certain additions to the netmap APIs. Finally in Section \ref{sec: } we present certain pseudo applications that involves operations such as compression and encryption to evaluate our parallel processing engine.

An important part of our system is the tasks related to packets such as routing, forwarding, etc., Since our design focuses more on the aspects of the parallel framework we utilize the click modular router for the packet processing tasks. Click modular router \cite{click} is a commonly used software router that includes a library of elements which is used in building flexible and reconfigurable routers. It offers a platform that is flexible for rapid development of newer network protocols and achieves good performance. An improved version known as SMP Click \cite{smpclick} also supports thread based parallel processing of network related tasks. Until recently click's performance in the user space was not as good as its kernel space equivalent. This is mainly due to its extensive use of polling driver in the kernel space. Netmap I/O \cite{netmap} provides a set of APIs that could directly access the packets from the NIC and it helped in the recent improvement of click's user space performance.

Click dynamically constructs routers in which the configuration file is passed as a parameter for the kernel module or user space binary. As a result most of the actual components are dynamically allocated and connected at run time. While this may be an advantage considering the alternate is to recompile the linux kernel, it does act as significant bottleneck in the case of user space. Moreover some of the restrictions on how the configuration can be parallelized limits the actual performance that could be achieved in combination with netmap. By integrating the ability to access click's elements within our parallel processing engine we were able to scale significantly better. This does require a limited amount of modifications to some of click's libraries and elements. In generating the actual code statically, compilers such as gnu c, etc., can perform optimizations such as loop fusions, in-lining, dead code optimization etc., which were not possible previously.

Our framework primarily focuses on user space and netmap APIs forms a crucial part in this.

The constant advancement in server hardware has made systems with large number of hardware threads[] readily available. In addition to this the presence of co-processors[], network processors[], Graphics Processing Units ( GPUs ) \& General Purpose GPUs ( GPGPUs ) has dramatically improved the capability of commodity hardware. These developments has paved way to a platform with immense computing potential which existing applications already take advantage of. Along with the improvements in the Multi-Core \& Many-Core processor architecture the memory and I/O speeds have also seen significant improvement. In particular the network subsystem has seen significant growth, in combination with the improvements to optical communication it has lead to 10GbE interfaces being quite common in enterprise servers. While 40GbE and 100GbE are available for backbone infrastructure but would eventually follow into commodity hardware.

Network applications represent those applications that operate on any of the seven network layers described in the OSI Model. \{describe more here\}

Several previous work have discussed the process of parallelizing network applications[][][] to enable it to run faster in such multi threaded systems. The development of Software routers[][][] is encouraged further because it costs only a fraction of the commodity hardware routers. Moreover the advancements in the network interfaces has made faster interfaces readily in combination with the multi core servers the performance  It has been shown that such routers can be used to process packets at line rate [][][] which has been a challenge on its own. Network applications including those that operate on few or all of the network layers have been designed with sequential processing as the target environment. The shift to multi core architectures in recent times has induced such applications to utilize parallel processing as a alternate means of computing. While this could lead to significant improvement in performance it also suffers from the same problems as any parallel application. These include scalability, load-balancing, data dependency and the presence of non parallel regions in the program apart from those that might arise due to that specific application. The scaling of software routers has been discussed extensively in RouteBricks[] which shows the problems suffered by software routers when they are scaled beyond a single system. One could argue that complexity of network applications such as Deep Packet Inspection ( DPI ), Packet Forwarding, etc., could be handled by a single many socket Multi-Core SMP system. However, as the complexity of network applications develops it would be inevitable but to move to a multi node environment in order to perform the computations and to handle more network bandwidth.

Large scale computing clusters has been used in high-performance scientific application to solve problems and applications of very large size. It uses a cluster of compute nodes and distributes the workload across them in order to solve it faster. These systems utilize parallel programs consisting of both threads and MPI to break massive problems into smaller workload that can be executed in parallel. In addition, the presence of GPUs and GPGPUs requires a mixture of parallelization constructs such as OpenCL, CUDA, etc., in order to achieve the best performance in the system. In this paper we discuss how some of the techniques used by these applications could be used to parallelize network applications.

Click Modular Software Router[] consists of extensive elements which are used to construct software routers. Click primarily targets packet processing applications especially those that operate on L4 - L7. \{elaborate more here\}

Discuss about what each sections contains here.....

Talk about click...\\

Describe about what would be the best way to parallelize using threads and MPI.. provide reference.. \\

\section{Design}

This section describes the design principles related to our methodology. The objective is to determine the requirements necessary to parallelize network applications. Our methods focus on ways to improve the click system in some cases even extend its functionality.

\subsection{Emphasize on user level execution environment}

The existing software routers are predominantly implemented in the kernel space for the ease of access to resources and can bypass the OS overhead incurred when data is accessed through the network subsystem. While this alternative provides better performance it also leads to concerns regarding stability and security of the system. Newer techniques such as Netmap[] allows direct access to network buffers which makes it possible to write high performance packet processing applications in userspace itself. Typically network applications which operate on L4 - L7 are developed in the kernel space in order to intercept the network packets before it flows through the OS. Whereas those applications that utilize L1 - L3 are present in the user space. The objective here is to completely switch to user space to solve any of the stability or security concerns. In addition, scenarios where the applications that operate across most or all the layers could be improved by combining the operations across the different layers in user space itself.

\subsection{Eliminate the overhead associated with creation of dynamic objects}

Click constructs application graphs based on dynamic creation of objects and passing the data packets across the virtual functions. Though this is a very elegant approach in creating the software router it does incur the penalty of the virtual function tables. The click-tool[] addresses this issue but it does not solve the problem entirely. The construction of the software router at run time is the primary reason behind it. This mainly arises due to the presence of kernel module which installs the router as a kernel module based on the input configuration. However in the user level it would not be necessary for a separate process to initialize the router instead we could generate the entire program statically including the necessary flow associated with the specific application. This not only eliminates the overhead associated but also opens up means to optimize the graph further.

\subsection{Optimize based on application graph}

>>>>

% *** SUBFIGURE PACKAGES ***
\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

\usepackage[font=small,center,belowskip=-10pt,aboveskip=5pt]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}

%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by their underlying communication system. In a typical implementation, the Operating System (OS) handles the majority of network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application. Therefore, a framework for computing and communicating in parallel is essential for improving performance.

We propose a high-level programming framework that integrates the packet processing operations with stream applications, and provides a common interface to execute both application and packet processing operations in parallel on multicore processors.

We utilize the ability of the network interface controller (NIC) to classify packets onto multiple hardware queues. This is combined with system software that allows direct access to these queues. We develop an interface in our framework for application processes to read or write in parallel. Our system integrates this with the application by a series of packet processing operations that transform the data from the application to packets suitable for transmission. Representing these tasks as stages in the application, we are able to parallelize them using techniques similar to those used for parallelizing regular computations.

All the communication operations in the framework are performed in userspace, bypassing the OS, to send and receive data directly over the network interface. By overcoming the OS bottleneck, and by parallelizing computations and packet operations together, we are able to boost the performance of stream applications substantially.

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
%Kevin - vertical spacing
\setlength{\textfloatsep}{15pt}
\section{Motivation}

Stream applications are designed in a manner where they operate on data that flows as a \textit{stream}. This data stream passes through different stages of computations, to be transformed from query to result \cite{Thies:2002:SLS:647478.727935} \cite{Halbwachs91thesynchronous} \cite{Stephens95asurvey}. This differs from a conventional computation model where data is stored and then processed. By eliminating the need to store data, and processing it on the go instead, it is possible for stream applications to process large amounts of data.

Stream applications such as file storage systems, database query management, data mining, etc., operating on vast amounts of data, are often distributed over multiple systems. Traditional stream applications target multicore processors for performing their computations in parallel. As multicore processors grow in power, so do stream applications, processing ever-increasing amounts of data. However, stream applications that operate on large amounts of data at a very high bandwidth can be limited by the communication stack of the OS kernel\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. With the continuing improvement of multicore and many-core architectures, it is becoming essential to improve the processing speeds of packet operations along with those of the application in order to scale performance.
\comment{
When performing packet processing operations for communication intensive stream applications, a major challenge is the high levels of performance required. These applications process large numbers of data streams in real time.  If the communication system cannot keep up with the rate at which packets arrive, then they must be dropped or remain unprocessed.
}

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using these features makes it possible to do much of the packet processing in parallel using the multiple cores commonly available on modern processors.

A major bottleneck for packet processing on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research uses alternate mechanisms to access packets from the network hardware directly to overcome this problem\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with particular system software that allows direct access to hardware-controlled packet queues, dramatically reducing overheads and improving performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on data between a source and sink in the form of a \textit{data stream}. Data in a stream must be processed sequentially, whereas those from separate streams can often be processed in parallel.

Modern network hardware typically has a mechanism for separating arriving packets into multiple queues, where packets that belong to the same flow are\comment{ guaranteed to be} placed in the same queue\cite{micro2008} \cite{intel2010}. We can utilize this feature to operate on multiple streams of data by processing each of the queues in parallel. However, building such applications is complex because it requires explicit parallel programming and interaction with low-level network interfaces. Research focuses on improving the performance of network applications by parallelizing the packet processing tasks on multicore architectures \cite{Dobrescu09routebricks:exploiting} \cite{Han:2010:PGS:1851275.1851207}. We parallelize both the application and packet operations to gain significant improvement in performance.

\section{Contribution}

In this paper we propose a high-level programming framework for building communication intensive stream applications. We focus on integrating the packet processing and application tasks into a single graph and using this to extract data parallelism. The developer describes their program as a graph of \textit{actors}. Each actor consumes one or more \textit{tokens} of data, which may be simple data such as integers, or more complex data such as packets, arrays, or objects and produces one or more tokens. The framework can be used to automatically map the application over multiple parallel cores, and interface with the \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} for accessing data through the network buffers.

The system can be divided into three main parts:

\begin{itemize}
\item A subsystem for interfacing with the network hardware through the \texttt{netmap} APIs to send and receive data from the application in user space without going through the operating system. (Section \ref{ppio})
\item Given a graph that represents the workload to be performed by the application, we construct and integrate the packet processing operations represented as a directed acyclic graph (DAG) to it. (Section \ref{ppe})
\item A run-time system that executes these tasks in parallel and explicitly schedules them on the multicore processor. (Section \ref{ppt})
\end{itemize}

% Even though we integrate the system with the application, the techniques for extracting parallelism from the application is not in scope of this paper. The application is considered to be in a streaming model and the tasks corresponding to the various functions are considered to be parallel in nature.

In this paper, we are concerned with applications that operate on data streams. Example operations of such an application are represented in the stream graph shown in figure \ref{fig1}. This sample application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression. Due to the nature of such stream applications, it is possible to distribute the component operations across different CPUs. We are particularly interested in a scenario where groups of networked servers are used to perform these kind of stream operations by distributing them evenly over different nodes.

\begin{figure}[ht]
\centering
\subfigure
{
	\includegraphics[width=3.5in]{ip-data-proc}
}

\subfigure
{
	\includegraphics[width=3in]{op-data-proc}
}
\caption{Sample stream application}
\label{fig1}
\end{figure}

The main challenge that arises in such a situation is how best to address the intensive computational and communication demands of such applications. Two important features enable us to tackle this challenge. Firstly, the operations of both the application and the network are parallel in nature. Secondly, multicore processors can be used to execute the application.

In our system, we use an efficient parallel processing framework that could be leveraged on multicore processors and multi queue network controllers to achieve efficient communication. The performance of multicore processors continues to dramatically improve the capability of commodity hardware, thus making it possible to meet the computational demand of such applications. In addition, multi queue network interface controllers, designed to handle multiple streams of data, help achieve the necessary communication performance.

An important part of our system is the tasks related to packet processing in protocols such as UDP, IP, etc., We utilize the modular router, Click, for these packet processing tasks. Click is a tool that includes a library of elements which is used in building flexible, re-configurable and modular software routers \cite{Kohler2000}. It offers a high-performance platform that is flexible for rapid development of newer network protocols. An improved version known as SMP Click supports thread-based parallel processing of network related tasks\cite{Chen:2001:FCP:647055.759948}. Until recently, Click's performance in the user space was inferior to its kernel space equivalent. This is mainly due to its extensive use of polling driver in the kernel space that was not possible in user space. Netmap I/O\cite{Rizzo:2012:RNI:2090147.2103536} provides a set of APIs that can directly access the packets from the NIC and was used to improve Click's user space performance. Although other other user space packet-access APIs could be used \cite{Rizzo:2012:RNI:2090147.2103536} \cite{1564468} \cite{Krasnyansky}, the simpler construction of \texttt{netmap} was more suitable to adapt to our framework.

Section \ref{dppl} represents the design principles that we follow to conceptualize our framework. Section \ref{imple} describes the implementation details used in the construction of the system. In section \ref{ppe}, we discuss the challenges in using a system consisting of execution engines to manage the work. The extent to which we extract parallelism ranges, not only to the packet tasks, but also to the low level access of data I/O to the network interface hardware through custom additions to the \texttt{netmap}. Finally in section \ref{app-const}, we present a simple stream application constructed from operations such as compression and encryption in order to evaluate our framework.

\section{Design Principles}
\label{dppl}

The design of a high-level programming framework that integrates packet processing tasks into the application requires overcoming several challenges. \comment{The entire system is driven by the extent to which parallelism can be extracted. Some of the techniques that can be applied to the packet processing tasks can be used for the tasks related to the application.} In this section, we describe these challenges and the principles that are used to address them. \comment{We focus on the methodologies related to the packet processing tasks and the parallel processing engine itself.}

\subsection{Userspace Packet I/O}

For a packet processing system to be integrated with an application, it is necessary to have an efficient mechanism for accessing the data from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. \comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

\subsubsection{Handling Multiple Hardware Queues}

The development of virtualization and improved network flow handling has led to hardware level multi queue support in NICs. During transmission, concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception, the NIC classifies each packet onto one of the receive queues in a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. This technique uses the header information or tuples to classify packets onto different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework for this reason. By offloading the classification of packets to the NIC, we remove any overhead associated with packet classification in the application. We can also use these hardware queues as a means of balancing the workload across different threads by controlling the number of queues that each process handles. \comment{Since different processes can access these transmit and receive queues, concurrent access to data is possible.}

\subsubsection{Flow based Packet Batching}

Packet batching is a common method used to boost the performance of network applications where several packets are combined and operated upon together, thus reducing the overhead of I/O operations. In our framework, batching is utilized in a form where the number of packets in a batch is varied based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms an important means by which we reduce the usage of system calls.

\subsubsection{Reduction of Per Packet System Calls}

System calls are required in order to synchronize data between the NIC hardware and the buffer memory used for intermediate storage of packets. Previous work describes the overhead associated with per-packet system calls and provides mechanisms for overcoming this overhead\cite{1564468}\cite{Han:2010:PGS:1851275.1851207}\cite{Rizzo:2012:RNI:2090147.2103536}.

We employ an \textit{adaptive scheme} described in section \ref{ppio} that utilizes this strategy of reducing per-packet system calls. Our framework balances the system calls based on the communication rate and application processing speed. The transmission and reception are handled differently to optimize the usage of system calls in each of the cases independently. During transmission, we issue system calls to synchronize the available buffer space based on the rate at which we transmit data. This prevents degradation in latency for larger packets and improves performance for smaller packets. During reception of packets, the system reads packets at the rate at which it is able to process them.

\subsection{Packet Processing Graph}

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph-based representation of packet processing tasks to determine the dependencies involved for the different stages and to determine the relation with the different tasks of the application. This graph contains the stages through which the data associated with the application is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application, and highlights those that have a data dependency on a specific task in the application. The graph also enables optimizations based on specific characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a graph representing User Datagram Protocol (UDP) packet generation.

\begin{figure}[ht]
\centering
\includegraphics[width=3.5in]{pack-op}
\caption{Simplified UDP packet construction}
\label{fig2}
\end{figure}

\subsection{Application Task Graph}

\comment{Application task graph represents the operations associated in performing the computations.} The subset of the application that we are interested in parallelizing is the stream processing component. In such components, the communication flow between tasks is well-defined. Both the packet processing and application tasks can be represent by a directed acyclic graph (DAG) similar to the packet processing graph. The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, optimizations in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with the packet processing tasks.

\subsection{Integration}

The sequence of tasks that can be executed in parallel is determined at this stage. The integration stage is used to combine the application graph and the packet processing graph into a larger graph that unifies and represents the overall operations. The dependency relations identified earlier are used in establishing links between the graphs. We determine the sequence of tasks that can be executed in parallel from both the application task graph and packet processing graph. The combined set of tasks is then used by the parallel processing engine.

\subsection{Parallel Processing Engine}

Two important aspects that the parallel processing engine must address is the scheduling of tasks and balancing the load across the available resources. The parallel processing engine consists of collaborating processes which are bound to specific CPUs and act as engines to execute the workload. Based on their requirements and the available resources, tasks are scheduled across these execution engines. \comment{In this manner, the workload can be distributed flexibly.}

\section{Implementation}
\label{imple}

\comment{Most extant stream applications depend on the OS for their communication tasks.}In commodity systems, the interaction with the OS for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications, it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic, the OS could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible, either by executing the application in the kernel space (which could compromise stability of the system), or by using an interface for accessing the NIC buffers from user space. The latter approach is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though these interfaces support user space packet access, they lack the ability to be invoked in a efficient manner by the application. More specifically, they do not have any way of optimizing the rate of processing, lacking, for example, the concept of flows for application tasks. The lack of support for interacting with these interfaces in order to read and write data in parallel from concurrent processes or threads adds to the problem. Also, since the application is now responsible for handling network packets, it has to perform the additional operations that would otherwise be carried out by the OS.

Our framework provides support for sending and receiving data in parallel over the NIC from user space and a common parallel execution engine that supports both application and packet processing tasks.

\subsection{Parallel Packet I/O}
\label{ppio}
The parallel packet I/O interface, when combined with the application tasks, provides concurrent communication for the application threads or processes over the network. To do this, two important requirements have to be met. Firstly, access to network data from user space should be provided, which can be read from and written to in parallel. Secondly, it should be possible to integrate the packet operations with the application by providing the necessary interfaces for performing these packet operations, thereby allowing for some flexibility in the choice of communication protocol.

\subsubsection{Userspace Packet Access}
Applications executed in kernel space for accessing hardware devices directly would lead to system instability. Instead, previous researchers \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access to network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in kernel space and user space to export the network hardware buffer's memory region. This enables the applications in user space to directly access the network hardware. It also prevents the application from accessing memory outside valid regions by checking the limits of the ring buffer. Packets can be written to the buffers when space is available and an \textit{ioctl} call to the driver enables the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in user space to check when new packets are available. On reading or writing packets, the available space is updated.

We leverage the availability of multiple hardware queues present in modern NICs to read and write data concurrently. \texttt{Netmap}, by default, supports assigning each hardware queue to an individual application process or thread. However, in order to allow for more efficient communication in the framework, we required the ability to assign more than one queue to a particular thread or process. Our modifications to \texttt{netmap}'s hardware queue assignment API allow our framework to balance the amount of packets handled by each process by varying the number of queues assigned to it.

During testing, it was found that it was essential to have control over the rate at which system calls were issued with respect to the amount of data being transferred. Frequent system calls hamper bandwidth and limit the peak performance of the application. To tackle this, we propose the following algorithms for reading and writing data from the hardware queues using the \texttt{Netmap} APIs. These algorithms are designed to provide the following features,

@@

\item Provide batching of packets based on available space on hardware rings
\item Reduce the usage of system calls such as \textit{ioctl} and \textit {poll}
\item Provide mechanisms for handling multiple queues through the application

@@

\comment{By minimizing the synchronizations with the hardware rings, we essentially limit the system calls issued and consequently improve the overall system performance.}

Algorithms \ref{s_pack} and \ref{r_pack} use a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has information about the hardware rings, updating it on a periodic basis. Once a fixed amount of space is determined, we update this on our local structure. We then continue to read or write data from the hardware ring, but update the actual limits and position on our local structure. When we have used up the space that was allocated, we update the hardware ring with the changes. In cases where the application is generating sparse traffic, it is possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive based on an application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue has proven to be effective in our experiments, but if an application requires a specific flow to be maintained, a different function here could provide that support. This would be helpful in cases where the data operations are more predictable, or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required, and effectively handling the multiple queues, we are able to achieve improved performance.

\setlength{\textfloatsep}{0pt}

\begin{algorithm}[h]
\caption{To send packets}
\label{s_pack}
\begin{algorithmic}[1]

\State \textbf{struct} $ring[ n ]$\textbf{\{}
\State \Comment{Stores information about hardware rings}
\State \Comment{ $ n \gets \ $tx\ rings\ assigned\ to\ thread}
\State \  \  $avb$, available space
\State \  \  $curr$, current position
\State \  \  $limit$, total space \ \textbf{\}}

\Function{send\_data\_mqueue}{Ring $f(n)$, Data $packet$}
\While{$!success$}
	\State $i \gets f( n ) $ \label{s_pack_param}
	\State $success = $ \texttt{SEND\_DATA}($ i, packet $)
\EndWhile
\EndFunction
\Function{send\_data}{Ring $i$, Data $packet$}
\If{ $ring[ i ].avb > 0$ }
	\State $hw\_ring[ i ].slot( ring[ i ].curr ) \gets packet$
	\State $ring[ n ].avb \gets ring[ i ].avb - 1$
	\State $ring[ i ].curr++$
	\If{ $ring[ i ].avb == 0$ }
		\State $hw\_ring[ i ].avb \ -=\ ring[ i ].limit$
		\State $hw\_ring[ i ].curr \gets ring[ i ].curr$
	\EndIf
	\State \textbf{return} $success$
\EndIf

\State \textbf{\textit{issue ioctl and poll}}
\State $ring[ i ].avb \gets hw\_ring[ i ].avb$
\State $ring[ i ].limit \gets hw\_ring[ i ].avb$
\EndFunction

\end{algorithmic}
\end{algorithm}

\subsubsection{Packet Processing Operations}

Packet processing operations consist of converting the raw data that the application provides into network packets suitable for transmission. Since the application tasks are represented as a set of operations on a data stream, we include these operations as stages through which the data flows through to get transformed into data packets. Integrating them with the application task graph makes it easier to extract parallelism later in the execution engine. Several user space libraries exist that perform packet related processing operations like libnids \cite{libnids}, lwip \cite{Dunkels01designand}, Click \cite{Kohler2000} etc. Since our application is targeted at a closed environment, we chose the Click modular router due to its flexibility and vast library of elements. Our current framework supports the UDP protocol but Click's modular infrastructure makes it easy to support more complex ones. \comment{Indeed, stream applications with more complex protocols would benefit even more from the framework due to the parallelizing of these complex tasks.}

Each of the elements in Click contain a virtual function which is called from the previous element it is attached to. It was necessary to overcome the limitation of Click's infrastructure where elements are constructed based on dynamically initialized objects and virtual functions. This was required due to the need to eliminate the overhead associated with the dynamic construction of the objects and virtual functions, and also to allow for optimizing the task graph along with these operations. To achieve these goals, our framework provides a simple support structure to allow Click elements to be imported and to be integrated with applications.
\begin{algorithm}[h]
\caption{To receive packets}
\label{r_pack}
\begin{algorithmic}[1]

\State \textbf{struct} $ring[ n ]$\textbf{\{}
\State \Comment{Stores information about hardware rings}
\State \Comment{ $ n \gets \ $rx\ rings\ assigned\ to\ thread}
\State \  \  $avb$, available packets
\State \  \  $curr$, current position
\State \  \  $limit$, total packets \ \textbf{\}}

\Function{receive\_data\_mqueue}{Ring $f(n)$}
\While{$!packet$}
	\State $i \gets f( n ) $ \label{r_pack_param}
	\State $packet =$ \texttt{RECEIVE\_DATA}(Ring $i$)
\EndWhile
\EndFunction

\Function{receive\_data}{Ring $i$}
\If{ $ring[ i ].avb > 0$ }
	\State $packet \gets hw\_ring[ i ].slot( ring[ i ].curr ) $
	\State $ring[ i ].avb \gets ring[ i ].avb - 1$
	\State $ring[ i ].curr++$
	\If{ $ring[ i ].avb < threshold$ } \label{r_pack_thres}
		\State $hw\_ring[ i ].avb -= ring[ i ].( limit - avb)$
		\State $hw\_ring[ i ].curr \gets ring[ i ].curr$
		\State $ring[ i ].avb \gets hw\_ring[ i ].avb$
		\State $ring[ i ].limit \gets hw\_ring[ i ].avb$
	\EndIf
	\State \textbf{return} $packet$
\EndIf

\State \textbf{\textit{issue ioctl and poll}}
\State $ring[ i ].avb \gets hw\_ring[ i ].avb$
\State $ring[ i ].limit \gets hw\_ring[ i ].avb$
\EndFunction

\end{algorithmic}
\end{algorithm}
\subsection{Parallel Processing Engine}

After integrating the packet operations by means of support from \texttt{Click}'s infrastructure and access to network hardware through \texttt{netmap}'s APIs, we integrate these operations with the application graph. The results is a complete task graph that does both the computation and communication operations. The next logical step is executing these tasks in parallel by exploiting data and task parallelism. Packet streams are inherently operable in a data parallel fashion i.e. it is possible to operate on two different streams simultaneously. Since the application tasks depend on the data packets, they could also be executed in parallel. So from the integrated graph constructed by combining the application and packet processing operations, it becomes possible to extract data parallelism by operating on different data streams simultaneously. By executing different components, representing distinct operations on the data streams, it is also possible to exploit task parallelism.

\subsubsection{Execution engine}
\label{ppe}
